---
title: "Melissa - Simulation study"
author:
- name: Andreas C. Kapourani
  affiliation: School of Informatics, University of Edinburgh, UK
  email: c.a.kapourani or kapouranis.andreas@gmail.com
- name: Guido Sanguinetti
  affiliation: School of Informatics, University of Edinburgh, UK
  email: G.Sanguinetti@ed.ac.uk
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r echo=FALSE, include=FALSE}
suppressPackageStartupMessages(library(BPRMeth))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(cowplot))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(library(grid))
suppressPackageStartupMessages(library(ROCR))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(RColorBrewer))
```

<!-- # Parse and filter data -->
```{r initial_params, echo=FALSE, include=FALSE}
# Data
io <- list()
io$script_dir <- "../"
io$data_dir   <- "../local-data/melissa/synthetic/imputation/"
io$cpg_dir    <- paste0(io$data_dir, "coverage/")
io$var_dir    <- paste0(io$data_dir, "dissimilarity/")
io$cell_dir    <- paste0(io$data_dir, "cells/")
io$K          <- 4
io$basis      <- 4
io$cpg_prcg   <- 0.4
io$cl_var     <- 0.5
io$data_prcg  <- 0.1
io$reg_prcg   <- 1
R.utils::sourceDirectory(paste0(io$script_dir, "lib/"), modifiedOnly = FALSE)
```


<!-- # Parse and filter data for Coverage analysis-->
```{r load_cpg_data, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}
# Different CpG coverages
io$cpg_analysis <- seq(0.1, 0.9, 0.1)
# Load synthetic data
cpg_data <- readRDS(paste0(io$cpg_dir, "raw/encode_data.rds")) 
# Load joint analysis results
dt_melissa_cpg_prcg <- readRDS(paste0(io$cpg_dir, "encode_melissa_K", io$K, "_rbf", io$basis, 
                                "_dataTrain", io$data_prcg, 
                                "_regionTrain", io$reg_prcg, 
                                "_clusterVar", io$cl_var, ".rds"))

# Load independent analysis results
indep_cpg_prcg <- readRDS(paste0(io$cpg_dir, "encode_indep_K", io$K, "_rbf", io$basis,
                                "_dataTrain", io$data_prcg,
                                "_regionTrain", io$reg_prcg,
                                "_clusterVar", io$cl_var, ".rds"))

# Load RF analysis results
rf_cpg_prcg <- readRDS(paste0(io$cpg_dir, "encode_rf_indep_K", io$K, "_rbf", io$basis,
                                "_dataTrain", io$data_prcg,
                                "_regionTrain", io$reg_prcg,
                                "_clusterVar", io$cl_var, ".rds"))

# Load RF analysis results
gmm_cpg_prcg <- readRDS(paste0(io$cpg_dir, "encode_gmm_K", io$K,
                                "_dataTrain", io$data_prcg,
                                "_regionTrain", io$reg_prcg,
                                "_clusterVar", io$cl_var, ".rds"))

N <- length(dt_melissa_cpg_prcg)    # Numer of replications
M <- length(io$cpg_analysis)  # Number of Cpg coverage thresholds

dt_cpg_analysis <- data.table(cpg_cov = numeric(), auc_melissa = numeric(),
                              auc_melissa_rate = numeric(), auc_indep_prof = numeric(),
                              auc_indep_rate = numeric(), auc_rf = numeric(), auc_gmm = numeric(),
                              f_melissa = numeric(),
                              f_melissa_rate = numeric(), f_indep_prof = numeric(),
                              f_indep_rate = numeric(), f_rf = numeric(), f_gmm = numeric(),
                              melissa_ari = numeric(), melissa_rate_ari = numeric(), gmm_ari = numeric(),
                              melissa_error = numeric(), melissa_rate_error = numeric(), gmm_error = numeric())

       
# Iterate over each replication of the experiment
for (i in 1:length(dt_melissa_cpg_prcg)) {     # Itrate over simulations
  for (m in 1:length(io$cpg_analysis)) {  # Iterate over CpG percentages
    # Create prediction objects
    melissa_pred <- prediction(dt_melissa_cpg_prcg[[i]]$eval_perf[[m]]$eval_prof$pred_obs,
                               dt_melissa_cpg_prcg[[i]]$eval_perf[[m]]$eval_prof$act_obs)
    melissa_rate_pred <- prediction(dt_melissa_cpg_prcg[[i]]$eval_perf[[m]]$eval_mean$pred_obs,
                               dt_melissa_cpg_prcg[[i]]$eval_perf[[m]]$eval_mean$act_obs)
    indep_prof_pred <- prediction(indep_cpg_prcg[[i]]$eval_perf[[m]]$eval_prof$pred_obs,
                               indep_cpg_prcg[[i]]$eval_perf[[m]]$eval_prof$act_obs)
    indep_rate_pred <- prediction(indep_cpg_prcg[[i]]$eval_perf[[m]]$eval_mean$pred_obs,
                               indep_cpg_prcg[[i]]$eval_perf[[m]]$eval_mean$act_obs)
    rf_pred <- prediction(rf_cpg_prcg[[i]]$eval_perf[[m]]$pred_obs, 
                          rf_cpg_prcg[[i]]$eval_perf[[m]]$act_obs)
    gmm_pred <- prediction(gmm_cpg_prcg[[i]]$eval_perf[[m]]$pred_obs, 
                           gmm_cpg_prcg[[i]]$eval_perf[[m]]$act_obs)
           
    # F-measure performance
    f_melissa <- performance(melissa_pred, "f")
    f_melissa_rate <- performance(melissa_rate_pred, "f")
    f_indep_prof <- performance(indep_prof_pred, "f")
    f_indep_rate <- performance(indep_rate_pred, "f")
    f_rf <- performance(rf_pred, "f")
    f_gmm <- performance(gmm_pred, "f")
    
    dt <- data.table(cpg_cov = io$cpg_analysis[m],
                     auc_melissa = performance(melissa_pred, "auc")@y.values[[1]],
                     auc_melissa_rate = performance(melissa_rate_pred, "auc")@y.values[[1]],
                     auc_indep_prof = performance(indep_prof_pred, "auc")@y.values[[1]],
                     auc_indep_rate = performance(indep_rate_pred, "auc")@y.values[[1]],
                     auc_rf = performance(rf_pred, "auc")@y.values[[1]],
                     auc_gmm = performance(gmm_pred, "auc")@y.values[[1]],
                     
                     f_melissa = f_melissa@y.values[[1]][min(which(f_melissa@x.values[[1]] <= 0.5))],
                     f_melissa_rate = f_melissa_rate@y.values[[1]][min(which(f_melissa_rate@x.values[[1]] <= 0.5))],
                     f_indep_prof = f_indep_prof@y.values[[1]][min(which(f_indep_prof@x.values[[1]] <= 0.5))],
                     f_indep_rate = f_indep_rate@y.values[[1]][min(which(f_indep_rate@x.values[[1]] <= 0.5))],
                     f_rf = f_rf@y.values[[1]][min(which(f_rf@x.values[[1]] <= 0.5))],
                     f_gmm = f_gmm@y.values[[1]][min(which(f_gmm@x.values[[1]] <= 0.5))],
                     melissa_ari = cluster_ari(cpg_data$synth_data[[i]]$C_true, dt_melissa_cpg_prcg[[i]]$melissa_prof[[m]]$r_nk),
                     melissa_rate_ari = cluster_ari(cpg_data$synth_data[[i]]$C_true, dt_melissa_cpg_prcg[[i]]$melissa_rate[[m]]$r_nk),
                     gmm_ari = cluster_ari(cpg_data$synth_data[[i]]$C_true, gmm_cpg_prcg[[i]]$gmm[[m]]$r_nk),
                     melissa_error = cluster_error(cpg_data$synth_data[[i]]$C_true, dt_melissa_cpg_prcg[[i]]$melissa_prof[[m]]$r_nk),
                     melissa_rate_error = cluster_error(cpg_data$synth_data[[i]]$C_true, dt_melissa_cpg_prcg[[i]]$melissa_rate[[m]]$r_nk),
                     gmm_error = cluster_error(cpg_data$synth_data[[i]]$C_true, gmm_cpg_prcg[[i]]$gmm[[m]]$r_nk)
                     )
    # Add results to final data.table
    dt_cpg_analysis <- rbind(dt_cpg_analysis, dt)
  }
}
rm(iter, dt, i, m, N, M, dt_melissa_cpg_prcg, indep_cpg_prcg, cpg_data, gmm_cpg_prcg, rf_cpg_prcg)
```

# AUC performance across CpG coverage 
```{r auc_plot_cpg, fig.wide=TRUE, echo=FALSE, include=TRUE, warning=FALSE, fig.width=12, fig.height=8}
set.seed(17)
auc_cpg_jitter <- copy(dt_cpg_analysis)
auc_cpg_jitter <- auc_cpg_jitter %>% .[, c("cpg_cov", "auc_melissa", "auc_melissa_rate", "auc_indep_prof", "auc_indep_rate", "auc_rf", "auc_gmm") ] %>% setnames(c("cpg_cov", "auc_melissa", "auc_melissa_rate", "auc_indep_prof", "auc_indep_rate", "auc_rf", "auc_gmm"), c("x", "Melissa", "Melissa Rate", "BPRMeth", "Rate", "RF", "GMM")) %>% .[, x := as.factor(x)] %>% melt(variable.name = "Model", value.name = "y") %>% .[, Model := factor(Model, levels = c("Melissa", "BPRMeth", "RF", "Melissa Rate", "Rate", "GMM"))]
p_auc_cpg_jitter <- auc_jitter_plot(auc_cpg_jitter, x_lab = "CpG coverage", y_lab = "AUC", title = "")
print(p_auc_cpg_jitter)
```

# F measure performance across CpG coverage 
```{r f_plot_cpg, fig.wide=TRUE, echo=FALSE, include=TRUE, warning=FALSE, fig.width=12, fig.height=8}
set.seed(17)
f_cpg_jitter <- copy(dt_cpg_analysis)
f_cpg_jitter <- f_cpg_jitter %>% .[, c("cpg_cov", "f_melissa", "f_melissa_rate", "f_indep_prof", "f_indep_rate", "f_rf", "f_gmm") ] %>% setnames(c("cpg_cov", "f_melissa", "f_melissa_rate", "f_indep_prof", "f_indep_rate", "f_rf", "f_gmm"), c("x", "Melissa", "Melissa Rate", "BPRMeth", "Rate", "RF", "GMM")) %>% .[, x := as.factor(x)] %>% melt(variable.name = "Model", value.name = "y") %>% .[, Model := factor(Model, levels = c("Melissa", "BPRMeth", "RF", "Melissa Rate", "Rate", "GMM"))]
p_f_cpg_jitter <- auc_jitter_plot(f_cpg_jitter, x_lab = "CpG coverage", y_lab = "F measure", title = "")
print(p_f_cpg_jitter)
```

<!-- # Parse and filter data. Cluster dissimilarity analysis-->
```{r load_diss_data, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}
# Different cluster dissimilarity values
io$cluster_var_analysis <- seq(0, 1, .1)
# Load joint analysis results
dt_melissa_cl_var <- readRDS(paste0(io$var_dir, "encode_melissa_K", io$K, "_rbf", io$basis, 
                                "_dataTrain", io$data_prcg, 
                                "_regionTrain", io$reg_prcg, 
                                "_cpgTrain", io$cpg_prcg, ".rds"))
# Load independent analysis results
indep_cl_var <- readRDS(paste0(io$var_dir, "encode_indep_K", io$K, "_rbf", io$basis, 
                                "_dataTrain", io$data_prcg, 
                                "_regionTrain", io$reg_prcg, 
                                "_cpgTrain", io$cpg_prcg, ".rds"))
# Load RF analysis results
rf_cl_var <- readRDS(paste0(io$var_dir, "encode_rf_indep_K", io$K, "_rbf", io$basis, 
                                "_dataTrain", io$data_prcg, 
                                "_regionTrain", io$reg_prcg, 
                                "_cpgTrain", io$cpg_prcg, ".rds"))

# Load RF analysis results
gmm_cl_var <- readRDS(paste0(io$var_dir, "encode_gmm_K", io$K,
                                "_dataTrain", io$data_prcg, 
                                "_regionTrain", io$reg_prcg, 
                                "_cpgTrain", io$cpg_prcg, ".rds"))

N <- length(dt_melissa_cl_var)             # Numer of replications
M <- length(io$cluster_var_analysis)  # Number of cell variability thresholds
dt_var_analysis <- data.table(cell_var = numeric(), auc_melissa = numeric(), 
                              auc_melissa_rate = numeric(), auc_indep_prof = numeric(), 
                              auc_indep_rate = numeric(), auc_rf = numeric(), auc_gmm = numeric(),
                              f_melissa = numeric(),
                              f_melissa_rate = numeric(), f_indep_prof = numeric(),
                              f_indep_rate = numeric(), f_rf = numeric(), f_gmm = numeric(),
                              melissa_ari = numeric(), melissa_rate_ari = numeric(), gmm_ari = numeric(),
                              melissa_error = numeric(), melissa_rate_error = numeric(), gmm_error = numeric())
# Iterate over each replication of the experiment
for (i in 1:length(dt_melissa_cl_var)) {
  for (m in 1:length(io$cluster_var_analysis)) {
    # Load synthetic data
    var_data <- readRDS(paste0(io$var_dir, "raw/data-sims/encode_data_", io$cluster_var_analysis[m], "_", i, ".rds")) 
    
    # Create prediction objects
    melissa_pred <- prediction(dt_melissa_cl_var[[i]]$eval_perf[[m]]$eval_prof$pred_obs,
                               dt_melissa_cl_var[[i]]$eval_perf[[m]]$eval_prof$act_obs)
    melissa_rate_pred <- prediction(dt_melissa_cl_var[[i]]$eval_perf[[m]]$eval_mean$pred_obs,
                               dt_melissa_cl_var[[i]]$eval_perf[[m]]$eval_mean$act_obs)
    indep_prof_pred <- prediction(indep_cl_var[[i]]$eval_perf[[m]]$eval_prof$pred_obs,
                               indep_cl_var[[i]]$eval_perf[[m]]$eval_prof$act_obs)
    indep_rate_pred <- prediction(indep_cl_var[[i]]$eval_perf[[m]]$eval_mean$pred_obs,
                               indep_cl_var[[i]]$eval_perf[[m]]$eval_mean$act_obs)
    rf_pred <- prediction(rf_cl_var[[i]]$eval_perf[[m]]$pred_obs, 
                          rf_cl_var[[i]]$eval_perf[[m]]$act_obs)
    gmm_pred <- prediction(gmm_cl_var[[i]]$eval_perf[[m]]$pred_obs, 
                          gmm_cl_var[[i]]$eval_perf[[m]]$act_obs)
           
    # F-measure performance
    f_melissa <- performance(melissa_pred, "f")
    f_melissa_rate <- performance(melissa_rate_pred, "f")
    f_indep_prof <- performance(indep_prof_pred, "f")
    f_indep_rate <- performance(indep_rate_pred, "f")
    f_rf <- performance(rf_pred, "f")
    f_gmm <- performance(gmm_pred, "f")
    
    dt <- data.table(cell_var = io$cluster_var_analysis[m], 
                     auc_melissa = performance(melissa_pred, "auc")@y.values[[1]],
                     auc_melissa_rate = performance(melissa_rate_pred, "auc")@y.values[[1]],
                     auc_indep_prof = performance(indep_prof_pred, "auc")@y.values[[1]],
                     auc_indep_rate = performance(indep_rate_pred, "auc")@y.values[[1]],
                     auc_rf = performance(rf_pred, "auc")@y.values[[1]],
                     auc_gmm = performance(gmm_pred, "auc")@y.values[[1]],
                     
                     f_melissa = f_melissa@y.values[[1]][min(which(f_melissa@x.values[[1]] <= 0.5))],
                     f_melissa_rate = f_melissa_rate@y.values[[1]][min(which(f_melissa_rate@x.values[[1]] <= 0.5))],
                     f_indep_prof = f_indep_prof@y.values[[1]][min(which(f_indep_prof@x.values[[1]] <= 0.5))],
                     f_indep_rate = f_indep_rate@y.values[[1]][min(which(f_indep_rate@x.values[[1]] <= 0.5))],
                     f_rf = f_rf@y.values[[1]][min(which(f_rf@x.values[[1]] <= 0.5))],
                     f_gmm = f_gmm@y.values[[1]][min(which(f_gmm@x.values[[1]] <= 0.5))],
                     
                     melissa_ari = cluster_ari(var_data$synth_data$C_true, dt_melissa_cl_var[[i]]$melissa_prof[[m]]$r_nk),
                     melissa_rate_ari = cluster_ari(var_data$synth_data$C_true, dt_melissa_cl_var[[i]]$melissa_rate[[m]]$r_nk),
                     gmm_ari = cluster_ari(var_data$synth_data$C_true, gmm_cl_var[[i]]$gmm[[m]]$r_nk),
                     melissa_error = cluster_error(var_data$synth_data$C_true, dt_melissa_cl_var[[i]]$melissa_prof[[m]]$r_nk),
                     melissa_rate_error = cluster_error(var_data$synth_data$C_true, dt_melissa_cl_var[[i]]$melissa_rate[[m]]$r_nk),
                     gmm_error = cluster_error(var_data$synth_data$C_true, gmm_cl_var[[i]]$gmm[[m]]$r_nk)
                     )
    # Add results to final data.table
    dt_var_analysis <- rbind(dt_var_analysis, dt)
  }
}
rm(iter, i, m, N, M, dt_melissa_cl_var, indep_cl_var, dt, var_data, gmm_cl_var, rf_cl_var)
```

# AUC performance across cluster dissimilarity 
```{r auc_plot_diss, fig.wide=TRUE, echo=FALSE, include=TRUE, warning=FALSE, fig.width=12, fig.height=8}
set.seed(17)
auc_var_jitter <- copy(dt_var_analysis)
auc_var_jitter <- auc_var_jitter %>% .[, c("cell_var", "auc_melissa", "auc_melissa_rate", "auc_indep_prof", "auc_indep_rate", "auc_rf", "auc_gmm") ] %>% setnames(c("cell_var", "auc_melissa", "auc_melissa_rate", "auc_indep_prof", "auc_indep_rate", "auc_rf", "auc_gmm"), c("x", "Melissa", "Melissa Rate", "BPRMeth", "Rate", "RF", "GMM")) %>% .[, x := as.factor(x)] %>% melt(variable.name = "Model", value.name = "y") %>% .[, Model := factor(Model, levels = c("Melissa", "BPRMeth", "RF", "Melissa Rate", "Rate", "GMM"))]
p_auc_var_jitter <- auc_jitter_plot(auc_var_jitter, x_lab = "Cluster dissimilarity", y_lab = "AUC", title = "")
print(p_auc_var_jitter)

# pdf(file = paste0("out/synthetic/auc-synth-var08.pdf"), width = 11, height = 6, useDingbats = FALSE)
# p_auc_var_jitter
# dev.off()
```

# F measure performance across cluster dissimilarity 
```{r f_plot_diss, fig.wide=TRUE, echo=FALSE, include=TRUE, warning=FALSE, fig.width=12, fig.height=8}
set.seed(17)
f_var_jitter <- copy(dt_var_analysis)
f_var_jitter <- f_var_jitter %>% .[, c("cell_var", "f_melissa", "f_melissa_rate", "f_indep_prof", "f_indep_rate", "f_rf", "f_gmm") ] %>% setnames(c("cell_var", "f_melissa", "f_melissa_rate", "f_indep_prof", "f_indep_rate", "f_rf", "f_gmm"), c("x", "Melissa", "Melissa Rate", "BPRMeth", "Rate", "RF", "GMM")) %>% .[, x := as.factor(x)] %>% melt(variable.name = "Model", value.name = "y") %>% .[, Model := factor(Model, levels = c("Melissa", "BPRMeth", "RF", "Melissa Rate", "Rate", "GMM"))]
p_f_var_jitter <- auc_jitter_plot(f_var_jitter, x_lab = "Cluster dissimilarity", y_lab = "F measure", title = "")
print(p_f_var_jitter)

# pdf(file = paste0("out/synthetic/f-synth-var08.pdf"), width = 11, height = 6, useDingbats = FALSE)
# p_f_var_jitter
# dev.off()
```




<!-- # Parse and filter data. Cells assayed analysis-->
```{r load_cell_assay_data, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}
# Different cluster dissimilarity values
io$cells_analysis <- c(25, 50, 75, 100, 125, 150, 175, 200, 225, 250) 
io$cpg_prcg <- 0.4
# Load joint analysis results
dt_melissa_cells <- readRDS(paste0(io$cell_dir, "encode_melissa_K", io$K, "_rbf", io$basis, 
                                "_dataTrain", io$data_prcg, 
                                "_regionTrain", io$reg_prcg, 
                                "_cpgTrain", io$cpg_prcg, ".rds"))
# Load independent analysis results
indep_cells <- readRDS(paste0(io$cell_dir, "encode_indep_K", io$K, "_rbf", io$basis, 
                                "_dataTrain", io$data_prcg, 
                                "_regionTrain", io$reg_prcg, 
                                "_cpgTrain", io$cpg_prcg, ".rds"))
# Load RF analysis results
rf_cells <- readRDS(paste0(io$cell_dir, "encode_rf_indep_K", io$K, "_rbf", io$basis, 
                                "_dataTrain", io$data_prcg, 
                                "_regionTrain", io$reg_prcg, 
                                "_cpgTrain", io$cpg_prcg, ".rds"))

# Load RF analysis results
gmm_cells <- readRDS(paste0(io$cell_dir, "encode_gmm_K", io$K,
                                "_dataTrain", io$data_prcg, 
                                "_regionTrain", io$reg_prcg, 
                                "_cpgTrain", io$cpg_prcg, ".rds"))

N <- length(dt_melissa_cells)             # Numer of replications
M <- length(io$cells_analysis)  # Number of cell variability thresholds
dt_cell_analysis <- data.table(cells = numeric(), auc_melissa = numeric(), 
                              auc_melissa_rate = numeric(), auc_indep_prof = numeric(), 
                              auc_indep_rate = numeric(), auc_rf = numeric(), auc_gmm = numeric(),
                              f_melissa = numeric(),
                              f_melissa_rate = numeric(), f_indep_prof = numeric(),
                              f_indep_rate = numeric(), f_rf = numeric(), f_gmm = numeric(),
                              melissa_ari = numeric(), melissa_rate_ari = numeric(), gmm_ari = numeric(),
                              melissa_error = numeric(), melissa_rate_error = numeric(), gmm_error = numeric())
# Iterate over each replication of the experiment
for (i in 1:length(dt_melissa_cells)) {
  for (m in 1:length(io$cells_analysis)) {
    # Load synthetic data
    cell_data <- readRDS(paste0(io$cell_dir, "raw/data-sims/encode_data_", io$cells_analysis[m], "_", i, ".rds")) 
    
    # Create prediction objects
    melissa_pred <- prediction(dt_melissa_cells[[i]]$eval_perf[[m]]$eval_prof$pred_obs,
                               dt_melissa_cells[[i]]$eval_perf[[m]]$eval_prof$act_obs)
    melissa_rate_pred <- prediction(dt_melissa_cells[[i]]$eval_perf[[m]]$eval_mean$pred_obs,
                               dt_melissa_cells[[i]]$eval_perf[[m]]$eval_mean$act_obs)
    indep_prof_pred <- prediction(indep_cells[[i]]$eval_perf[[m]]$eval_prof$pred_obs,
                               indep_cells[[i]]$eval_perf[[m]]$eval_prof$act_obs)
    indep_rate_pred <- prediction(indep_cells[[i]]$eval_perf[[m]]$eval_mean$pred_obs,
                               indep_cells[[i]]$eval_perf[[m]]$eval_mean$act_obs)
    rf_pred <- prediction(rf_cells[[i]]$eval_perf[[m]]$pred_obs, 
                          rf_cells[[i]]$eval_perf[[m]]$act_obs)
    gmm_pred <- prediction(gmm_cells[[i]]$eval_perf[[m]]$pred_obs, 
                          gmm_cells[[i]]$eval_perf[[m]]$act_obs)
           
    # F-measure performance
    f_melissa <- performance(melissa_pred, "f")
    f_melissa_rate <- performance(melissa_rate_pred, "f")
    f_indep_prof <- performance(indep_prof_pred, "f")
    f_indep_rate <- performance(indep_rate_pred, "f")
    f_rf <- performance(rf_pred, "f")
    f_gmm <- performance(gmm_pred, "f")
    
    dt <- data.table(cells = io$cells_analysis[m], 
                     auc_melissa = performance(melissa_pred, "auc")@y.values[[1]],
                     auc_melissa_rate = performance(melissa_rate_pred, "auc")@y.values[[1]],
                     auc_indep_prof = performance(indep_prof_pred, "auc")@y.values[[1]],
                     auc_indep_rate = performance(indep_rate_pred, "auc")@y.values[[1]],
                     auc_rf = performance(rf_pred, "auc")@y.values[[1]],
                     auc_gmm = performance(gmm_pred, "auc")@y.values[[1]],
                     
                     f_melissa = f_melissa@y.values[[1]][min(which(f_melissa@x.values[[1]] <= 0.5))],
                     f_melissa_rate = f_melissa_rate@y.values[[1]][min(which(f_melissa_rate@x.values[[1]] <= 0.5))],
                     f_indep_prof = f_indep_prof@y.values[[1]][min(which(f_indep_prof@x.values[[1]] <= 0.5))],
                     f_indep_rate = f_indep_rate@y.values[[1]][min(which(f_indep_rate@x.values[[1]] <= 0.5))],
                     f_rf = f_rf@y.values[[1]][min(which(f_rf@x.values[[1]] <= 0.5))],
                     f_gmm = f_gmm@y.values[[1]][min(which(f_gmm@x.values[[1]] <= 0.5))],
                     
                     melissa_ari = cluster_ari(cell_data$synth_data$C_true, dt_melissa_cells[[i]]$melissa_prof[[m]]$r_nk),
                     melissa_rate_ari = cluster_ari(cell_data$synth_data$C_true, dt_melissa_cells[[i]]$melissa_rate[[m]]$r_nk),
                     gmm_ari = cluster_ari(cell_data$synth_data$C_true, gmm_cells[[i]]$gmm[[m]]$r_nk),
                     melissa_error = cluster_error(cell_data$synth_data$C_true, dt_melissa_cells[[i]]$melissa_prof[[m]]$r_nk),
                     melissa_rate_error = cluster_error(cell_data$synth_data$C_true, dt_melissa_cells[[i]]$melissa_rate[[m]]$r_nk),
                     gmm_error = cluster_error(cell_data$synth_data$C_true, gmm_cells[[i]]$gmm[[m]]$r_nk)
                     )
    # Add results to final data.table
    dt_cell_analysis <- rbind(dt_cell_analysis, dt)
  }
}
rm(iter, i, m, N, M, dt_melissa_cells, indep_cells, rf_cells, gmm_cells, dt, cell_data)
```


# AUC performance across cells assayed 
```{r auc_plot_cells, fig.wide=TRUE, echo=FALSE, include=TRUE, warning=FALSE, fig.width=12, fig.height=8}
set.seed(17)
auc_cell_jitter <- copy(dt_cell_analysis)
auc_cell_jitter <- auc_cell_jitter %>% .[, c("cells", "auc_melissa", "auc_melissa_rate", "auc_indep_prof", "auc_indep_rate", "auc_rf", "auc_gmm") ] %>% setnames(c("cells", "auc_melissa", "auc_melissa_rate", "auc_indep_prof", "auc_indep_rate", "auc_rf", "auc_gmm"), c("x", "Melissa", "Melissa Rate", "BPRMeth", "Rate", "RF", "GMM")) %>% .[, x := as.factor(x)] %>% melt(variable.name = "Model", value.name = "y") %>% .[, Model := factor(Model, levels = c("Melissa", "BPRMeth", "RF", "Melissa Rate", "Rate", "GMM"))]
p_auc_cell_jitter <- auc_jitter_plot(auc_cell_jitter, x_lab = "Cells assayed", y_lab = "AUC", title = "")
print(p_auc_cell_jitter)

# pdf(file = paste0("out/synthetic/auc-cells.pdf"), width = 11, height = 6, useDingbats = FALSE)
# p_auc_cell_jitter
# dev.off()
```


# F measure performance across cells assayed 
```{r f_plot_diss, fig.wide=TRUE, echo=FALSE, include=TRUE, warning=FALSE, fig.width=12, fig.height=8}
set.seed(17)
f_cell_jitter <- copy(dt_cell_analysis)
f_cell_jitter <- f_cell_jitter %>% .[, c("cells", "f_melissa", "f_melissa_rate", "f_indep_prof", "f_indep_rate", "f_rf", "f_gmm") ] %>% setnames(c("cells", "f_melissa", "f_melissa_rate", "f_indep_prof", "f_indep_rate", "f_rf", "f_gmm"), c("x", "Melissa", "Melissa Rate", "BPRMeth", "Rate", "RF", "GMM")) %>% .[, x := as.factor(x)] %>% melt(variable.name = "Model", value.name = "y") %>% .[, Model := factor(Model, levels = c("Melissa", "BPRMeth", "RF", "Melissa Rate", "Rate", "GMM"))]
p_f_cell_jitter <- auc_jitter_plot(f_cell_jitter, x_lab = "Cluster dissimilarity", y_lab = "F measure", title = "")
print(p_f_cell_jitter)

# pdf(file = paste0("out/synthetic/f-synth-var08.pdf"), width = 11, height = 6, useDingbats = FALSE)
# p_f_cell_jitter
# dev.off()
```


# ARI performance CpG coverage
```{r ari_plot_cpg, fig.wide=TRUE, echo=FALSE, include=TRUE, warning=FALSE, fig.width=10, fig.height=7}
dt_ari_cpg_jitter <- copy(dt_cpg_analysis)
dt_ari_cpg_jitter <- dt_ari_cpg_jitter %>% .[, c("cpg_cov", "melissa_ari", "melissa_rate_ari", "gmm_ari")] %>% setnames(c("cpg_cov", "melissa_ari", "melissa_rate_ari", "gmm_ari"), c("x", "Melissa", "Melissa Rate", "GMM")) %>% .[, x := as.factor(x)] %>% melt(variable.name = "Model", value.name = "y")
p_ari_cpg_jitter <- ari_jitter_plot(dt_ari_cpg_jitter, x_lab = "CpG coverage", y_lab = "ARI", title = "")
print(p_ari_cpg_jitter)
```


# ARI performance cluster dissimilarity
```{r ari_plot_diss, fig.wide=TRUE, echo=FALSE, include=TRUE, warning=FALSE, fig.width=10, fig.height=7}
dt_ari_var_jitter <- copy(dt_var_analysis)
dt_ari_var_jitter <- dt_ari_var_jitter %>% .[, c("cell_var", "melissa_ari", "melissa_rate_ari", "gmm_ari")] %>% setnames(c("cell_var", "melissa_ari", "melissa_rate_ari", "gmm_ari"), c("x", "Melissa", "Melissa Rate", "GMM")) %>% .[, x := as.factor(x)] %>% melt(variable.name = "Model", value.name = "y")
p_ari_var_jitter <- ari_jitter_plot(dt_ari_var_jitter, x_lab = "Cluster dissimilarity", y_lab = "ARI", title = "")
print(p_ari_var_jitter)

# pdf(file = paste0("out/synthetic/gmm/ari-synth.pdf"), width = 10, height = 5, useDingbats = FALSE)
# p_ari_var_jitter
# dev.off()
```


# ARI performance cells assayed
```{r ari_plot_diss, fig.wide=TRUE, echo=FALSE, include=TRUE, warning=FALSE, fig.width=10, fig.height=7}
dt_ari_cell_jitter <- copy(dt_cell_analysis)
dt_ari_cell_jitter <- dt_ari_cell_jitter %>% .[, c("cells", "melissa_ari", "melissa_rate_ari", "gmm_ari")] %>% setnames(c("cells", "melissa_ari", "melissa_rate_ari", "gmm_ari"), c("x", "Melissa", "Melissa Rate", "GMM")) %>% .[, x := as.factor(x)] %>% melt(variable.name = "Model", value.name = "y")
p_ari_cell_jitter <- ari_jitter_plot(dt_ari_cell_jitter, x_lab = "Cells assayed", y_lab = "ARI", title = "")
print(p_ari_cell_jitter)

# pdf(file = paste0("out/synthetic/gmm/ari-synth-cells.pdf"), width = 10, height = 5, useDingbats = FALSE)
# p_ari_cell_jitter
# dev.off()
```


# Model selection
```{r model_selection, fig.wide=TRUE, echo=FALSE, include=TRUE, warning=FALSE, fig.width=10, fig.height=7}
# Data
io <- list()
io$script_dir    <- "../"
io$model_sel_dir <- "../local-data/melissa/synthetic/model-selection/"
io$K <- 10
# Different CpG coverages
io$cluster_var_analysis <- seq(0, 1, .1)
# Load joint analysis results
broad_model <- readRDS(paste0(io$model_sel_dir, "encode_broad_model_selection_K", io$K, ".rds"))
strict_model <- readRDS(paste0(io$model_sel_dir, "encode_strict_model_selection_K", io$K, ".rds"))

dt_model_analysis <- data.table(cell_var = numeric(), 
                              broad_K = numeric(), 
                              strict_K = numeric())
for (i in 1:5) {
  for (m in 1:length(io$cluster_var_analysis)) {
      dt <- data.table(cell_var = io$cluster_var_analysis[m], 
                       broad_K = length(which(broad_model[[i]]$bpr_prof_fit[[m]]$delta > 6)),
                       strict_K = length(which(strict_model[[i]]$bpr_prof_fit[[m]]$delta > 6)))
    dt_model_analysis <- rbind(dt_model_analysis, dt)
  }
}
rm(dt, i, m, broad_model, strict_model)

dt_model_jitter <- copy(dt_model_analysis)
dt_model_jitter <- dt_model_jitter %>% setnames(c("cell_var", "broad_K", "strict_K"), 
                                                c("x", "Broad Prior", "Strict Prior")) %>% 
    .[, x := as.factor(x)] %>% melt(variable.name = "Model", value.name = "y")
p_model_jitter <- ari_jitter_plot(dt_model_jitter, x_lab = "Cluster dissimilarity", 
                                  y_lab = "Clusters K", title = "")
print(p_model_jitter)
```


# Model efficicency
```{r model_efficiency, fig.wide=TRUE, warning=FALSE, echo=FALSE, include=TRUE, fig.width=10, fig.height=7}
# Data
library(microbenchmark)
io <- list()
io$script_dir    <- "../"
io$model_eff_dir <- "../local-data/melissa/synthetic/model-efficiency/"
io$K <- 3
io$M <- 200
io$N <- c(50, 100, 200, 500, 1000, 2000)

io$iter <- seq(1, 5)
dt_efficiency_analysis <- data.table(cells = numeric(), min_gibbs = numeric(), min_vb = numeric())
for (i in io$iter) {
    for (n in io$N) {
        eff_dt <- readRDS(paste0(io$model_eff_dir, "model_efficiency_K", io$K, "_M", 
                                 io$M, "_N", n, "_", i, ".rds"))[[1]]
        dt <- data.table(cells = n, min_gibbs = (eff_dt$time[which(eff_dt[, "expr"] == "gibbs")]) / 10^9 / 60 / 60,
                         min_vb = (eff_dt$time[which(eff_dt[, "expr"] == "vb")]) / 10^9 / 60 / 60 )
    dt_efficiency_analysis <- rbind(dt_efficiency_analysis, dt)
    }
}
rm(i, n, eff_dt)

dt_eff_jitter <- copy(dt_efficiency_analysis)
dt_eff_jitter <- dt_eff_jitter %>% setnames(c("cells", "min_gibbs", "min_vb"), c("x", "Gibbs", "VB")) %>% 
    .[, x := as.factor(x)] %>% melt(variable.name = "Model", value.name = "y") %>% 
    .[, Model := factor(Model, levels = c("VB", "Gibbs"))]
p_eff_jitter <- eff_jitter_plot(dt_eff_jitter, x_lab = "Cells", y_lab = "Hours", title = "")
print(p_eff_jitter)
```


# Joint plot AUC ENCODE synthetic
```{r joint_plot_auc, fig.wide=TRUE, eval=TRUE, echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.height=7, fig.width=16}
## AUC plot
p_auc_cpg_jitter <- auc_jitter_plot(auc_cpg_jitter, x_lab = "CpG coverage", y_lab = "AUC", title = "") +
    theme(legend.position = "none") + scale_y_continuous(limits = c(0.5, .9), breaks = pretty_breaks(n = 6))
p_auc_cell_jitter <- auc_jitter_plot(auc_cell_jitter, x_lab = "Cells assayed", y_lab = "AUC", title = "") +
    theme(legend.position = "right") + scale_y_continuous(limits = c(0.5, .9), breaks = pretty_breaks(n = 6))
# p_auc_var_jitter <- auc_jitter_plot(auc_var_jitter, x_lab = "Cluster dissimilarity", y_lab = "AUC", title = "") +
#     theme(legend.position = "right") + scale_y_continuous(limits = c(0.5, .9), breaks = pretty_breaks(n = 6))

final_fig_auc <- plot_grid(p_auc_cpg_jitter, p_auc_cell_jitter, labels = c("a", "b"), 
                           label_size = 25, ncol = 2, nrow = 1, rel_widths = c(1, 1.5))
print(final_fig_auc)

pdf(file = paste0("out/synthetic/auc-cpg-cells.pdf"), width = 17, height = 6, useDingbats = FALSE)
final_fig_auc
dev.off()
```


# Joint plot F-measure ENCODE synthetic
```{r joint_plot_f, fig.wide=TRUE, eval=TRUE, echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.height=7, fig.width=16}
## AUC plot
p_f_cpg_jitter <- auc_jitter_plot(f_cpg_jitter, x_lab = "CpG coverage", y_lab = "F-measure", title = "") +
    theme(legend.position = "none") + scale_y_continuous(limits = c(0.3, .8), breaks = pretty_breaks(n = 6))
p_f_cell_jitter <- auc_jitter_plot(f_cell_jitter, x_lab = "Cells assayed", y_lab = "F-measure", title = "") +
    theme(legend.position = "right") + scale_y_continuous(limits = c(0.3, .8), breaks = pretty_breaks(n = 6))

final_fig_f <- plot_grid(p_f_cpg_jitter, p_f_cell_jitter, labels = c("a", "b"), 
                           label_size = 25, ncol = 2, nrow = 1, rel_widths = c(1, 1.5))
print(final_fig_f)

pdf(file = paste0("out/synthetic/f-cpg-cells.pdf"), width = 17, height = 6, useDingbats = FALSE)
final_fig_f
dev.off()
```


# Joint plot ENCODE synthetic, cluster dissimilarity
```{r joint_plot_dissimilarity, fig.wide=TRUE, eval=TRUE, echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.height=7, fig.width=16}
## AUC plot
p_auc_var_jitter <- auc_jitter_plot(auc_var_jitter, x_lab = "Cluster dissimilarity", y_lab = "AUC", title = "") +
    theme(legend.position = "none") + scale_y_continuous(limits = c(0.5, .9), breaks = pretty_breaks(n = 6))
p_f_var_jitter <- auc_jitter_plot(f_var_jitter, x_lab = "Cluster dissimilarity", y_lab = "F-measure", title = "") +
    theme(legend.position = "right") + scale_y_continuous(limits = c(0.3, .8), breaks = pretty_breaks(n = 6))

final_fig_var <- plot_grid(p_auc_var_jitter, p_f_var_jitter, labels = c("a", "b"), 
                           label_size = 25, ncol = 2, nrow = 1, rel_widths = c(1, 1.45))
print(final_fig_var)

pdf(file = paste0("out/synthetic/cl-var.pdf"), width = 17, height = 6, useDingbats = FALSE)
final_fig_var
dev.off()
```


# Joint plot performance ENCODE synthetic
```{r joint_plot_analysis, fig.wide=TRUE, eval=TRUE, echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.height=10, fig.width=16}
# Joint plot of ARI, model selection and model efficiency
p_ari_var_jitter <- ari_jitter_plot(dt_ari_var_jitter, x_lab = "Cluster dissimilarity", y_lab = "ARI", title = "Cluster performance") +
    theme(legend.position = "none")  +
    scale_x_discrete(breaks = pretty_breaks(n = 6))
p_ari_cpg_jitter <- ari_jitter_plot(dt_ari_cpg_jitter, x_lab = "CpG coverage", y_lab = "ARI", title = "Cluster performance") +
    theme(legend.position = c(0.55, 0.23))  +
    scale_x_discrete(breaks = pretty_breaks(n = 5))
p_ari_cell_jitter <- ari_jitter_plot(dt_ari_cell_jitter, x_lab = "Cells assayed", y_lab = "ARI", title = "Cluster performance") +
    theme(legend.position = "none")  #+
    #scale_x_discrete(breaks = pretty_breaks(n = 6))
p_model_jitter <- ari_jitter_plot(dt_model_jitter, x_lab = "Cluster dissimilarity", y_lab = "Clusters K", title = "Model Selection") +
    theme(legend.position = c(0.55, 0.23)) +
    scale_x_discrete(breaks = pretty_breaks(n = 6))

top_ari_analysis <- plot_grid(p_ari_cpg_jitter, p_ari_var_jitter,  labels = c("a", "b"), 
                              label_size = 25, ncol = 2, nrow = 1, rel_widths = c(1, 1))
bot_perf_analysis <- plot_grid(p_ari_cell_jitter, p_model_jitter,  labels = c("c", "d"), 
                               label_size = 25, ncol = 2, nrow = 1, rel_widths = c(1, 1))

final_encode_analysis <- plot_grid(top_ari_analysis, bot_perf_analysis, ncol = 1, nrow = 2, rel_widths = c(1, 1))
print(final_encode_analysis)

pdf(file = paste0("out/synthetic/perf-synth.pdf"), width = 14, height = 10, useDingbats = FALSE)
final_encode_analysis
dev.off()
```


# Joint plot cells assayed - imputation and clustering
```{r joint_plot_cells_imputation_clustering, fig.wide=TRUE, eval=TRUE, echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.height=6, fig.width=16}
# Joint plot of ARI, model selection and model efficiency
p_ari_cell_jitter <- ari_jitter_plot(dt_ari_cell_jitter, x_lab = "Cells assayed", y_lab = "ARI", title = "Cluster performance") +
    theme(legend.position = c(0.55, 0.23))
    #scale_x_discrete(breaks = pretty_breaks(n = 6))

final_fig_cell_imp_cl <- plot_grid(p_auc_cell_jitter, p_ari_cell_jitter, labels = c("a", "b"), 
                           label_size = 25, ncol = 2, nrow = 1, rel_widths = c(1.3, 1))
print(final_fig_cell_imp_cl)

pdf(file = paste0("out/synthetic/auc-ari-cells.pdf"), width = 17, height = 6, useDingbats = FALSE)
final_fig_cell_imp_cl
dev.off()
```


# Efficiency plot performance ENCODE synthetic
```{r efficiency_plot_analysis, fig.wide=TRUE, eval=TRUE, echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.height=14, fig.width=16}
p_eff_jitter <- eff_jitter_plot(dt_eff_jitter, x_lab = "Cells", y_lab = "Hours", title = "Model Cost") +
    theme(legend.position = c(0.06, 0.8))
print(p_eff_jitter)
pdf(file = paste0("out/synthetic/efficiency.pdf"), width = 10, height = 5, useDingbats = FALSE)
p_eff_jitter
dev.off()
```
